---
title: "AddNeuroMed dataset - Stratified by Sex"
author: "Tiago C. Silva, Wei Zhang, Lanyu Zhang, Lily Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```

Delete the remove smoking part and re-run the code.

# Setup

```{R, message = FALSE, results = 'hide'}
library(dplyr)
library(SummarizedExperiment)
library(ExperimentHub)
cohort <- "AddNeuroMed"
dir.base <- "~/TBL Dropbox/Wei Zhang/AD-meta-analysis-blood-samples/"
dir.base2 <- "~/TBL Dropbox/Wei Zhang/AD-meta-analysis-blood-samples-bySex/"
dir.data <- file.path(dir.base,"datasets/",cohort,"/") 
dir.data.aux <- file.path(dir.base,"datasets/Aux/") 
dir.data.raw <- file.path(dir.data,"/step1_download/") 
dir.data.processed <- file.path(dir.data,"/step2_processed/withSmoke/") 
dir.data.pca <- file.path(dir.data,"/step3_pca_filtering/withSmoke/") 
dir.results <- file.path(dir.base2,"analysis_results",cohort,'withSmoke/')
dir.results.glm <- file.path(dir.results,"/Logistic_regression_model/")
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```


# Data retrieval

```{R, load_rgset_bs_filtered_min_65, include = FALSE, eval = T}
load(file.path(dir.data.raw,"addNeuroMed_se_non_normalized.rda"))
nb.samples.male <- sum(addNeuroMed$Sex.ch1 == "Male")
nb.samples.female <- sum(addNeuroMed$Sex.ch1 == "Female")
nb.probes.before.norm <- nrow(addNeuroMed)

addNeuroMed <- addNeuroMed[,addNeuroMed$age.65.exclusion.ch1 == "Included"]
nb.samples.age.65.exclusion <- ncol(addNeuroMed)
addNeuroMed$age.ch1 <-  as.numeric(addNeuroMed$age.ch1)
nb.samples.age.65.exclusion.male <- sum(addNeuroMed$Sex.ch1 == "Male")
nb.samples.age.65.exclusion.female <- sum(addNeuroMed$Sex.ch1 == "Female")

```

```{R, load_bs_filtered_min_65, include = FALSE, eval = T}
addNeuroMed.se <- readRDS(file.path(dir.data.pca, "addNeuroMed_QNBMIQ_PCfiltered.RDS"))

nb.samples.after.pca.female <-  sum(addNeuroMed.se$Sex.ch1 == "Female")
nb.samples.after.pca.female.ad.cn <-  sum(addNeuroMed.se$Sex.ch1 == "Female" & addNeuroMed.se$disease.state.ch1 != "mild cognitive impairment")

nb.samples.after.pca.male <- sum(addNeuroMed.se$Sex.ch1 == "Male")
nb.samples.after.pca.male.ad.cn <-  sum(addNeuroMed.se$Sex.ch1 == "Male" & addNeuroMed.se$disease.state.ch1 != "mild cognitive impairment")

```

## Probes QC

1. keep only probes that start with "cg"
2. drop probes where SNP with MAF >= 0.01 was present in the last 5 bp of the probe.

```{R, eval = FALSE}
library(minfi)
library(wateRmelon)
library(RPMM)
library(DMRcate)
library(sesameData)
```

```{R, eval = FALSE}
eh = query(ExperimentHub(),"sesameData")
annotation <- eh[['EH3674']] %>% as.data.frame %>% dplyr::filter(seqnames %in% c("chrX","chrY"))
betaSetXY <- addNeuroMed[rownames(addNeuroMed) %in% rownames(annotation), ]
nb.chrXY.probes <- nrow(betaSetXY)

# nrow(betaSetXY)
# 9676

betaSetXY <- betaSetXY[grep("cg",rownames(betaSetXY)),]
nb.chrXY.cg.probes <- nrow(betaSetXY)

# nrow(betaSetXY)
# 9623

gc()

# snapshotDate(): 2021-03-30
betaSetXY <- rmSNPandCH(
  object = assay(betaSetXY),
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = TRUE,
  rmXY = FALSE
)
```

```{R, include = F}
# dim(betaSetXY)
#  9546  284
```




## Merge DNA methylation matrix on autosome with the one on chrX, chrY

```{R, eval = FALSE}
betaSetXY <- betaSetXY[,colnames(addNeuroMed.se)]
identical(colnames(addNeuroMed.se), colnames(betaSetXY))
betaQN_BMIQ <- rbind(assay(addNeuroMed.se), betaSetXY)
table(base::duplicated(row.names(betaQN_BMIQ)))
clinical.filtered.min.age_at_visit.65 <- colData(addNeuroMed.se)
# 403939
save(
  betaQN_BMIQ,
  clinical.filtered.min.age_at_visit.65,
  file = file.path(dir.data.processed,"RGSet_bs_filtered_min_age_at_visit_65_with_XY.rda")
)
```

### Summarize Experiment

```{R, eval = FALSE}
betaQN_BMIQ <- betaQN_BMIQ[
  rownames(betaQN_BMIQ) %in% rownames(IlluminaHumanMethylation450kanno.ilmn12.hg19::Locations),
]

assay <- S4Vectors::SimpleList("DNAm" = betaQN_BMIQ)

samples.info <- colData(addNeuroMed.se)


rowData <- IlluminaHumanMethylation450kanno.ilmn12.hg19::Locations[rownames(betaQN_BMIQ),] %>%
  GenomicRanges::makeGRangesFromDataFrame(start.field = "pos",end.field = "pos")


addNeuroMed.se <- SummarizedExperiment::SummarizedExperiment(
  assays = assay,
  rowData = rowData,
  metadata = list("genome" = "hg19"),
  colData = samples.info
)

saveRDS(addNeuroMed.se, file = file.path(dir.data.processed,"addNeuroMed_se_withXY.rds")) # 403939 280
readr::write_csv(samples.info %>% as.data.frame, file = file.path(dir.data.processed,"samples.info_withXY.csv"))
```


# Summary after QC steps


```{R, eval = F, echo = F}
addNeuroMed.se <- readRDS(file.path(dir.data.processed,"addNeuroMed_se_withXY.rds"))
addNeuroMed.se <- addNeuroMed.se[,addNeuroMed.se$disease.state.ch1 %in% c("Alzheimer's disease" ,"control")]
pheno_df <- colData(addNeuroMed.se) %>% as.data.frame
nb.samples.before.pca.male <- sum(pheno_df$Sex.ch1 == "Female")
nb.samples.before.pca.female <- sum(pheno_df$Sex.ch1 == "Male")
```

## Data and metadata
```{R}
addNeuroMed.se <- readRDS(file.path(dir.data.processed,"addNeuroMed_se_withXY.rds"))
addNeuroMed.se <- addNeuroMed.se[,addNeuroMed.se$disease.state.ch1 %in% c("Alzheimer's disease" ,"control")]
nb.samples.after.pca <- ncol(addNeuroMed.se)
nb.probes.after.norm <- nrow(addNeuroMed.se)
```



```{R}
pheno.cn.ad <- colData(addNeuroMed.se)
pheno.cn.ad.female <- pheno.cn.ad[pheno.cn.ad$Sex.ch1 == "Female" & pheno.cn.ad$disease.state.ch1 %in% c("Alzheimer's disease" ,"control"),]
nb.samples.after.pca.female <- nrow(pheno.cn.ad.female)
```

```{R}
pheno.cn.ad.male <- pheno.cn.ad[pheno.cn.ad$Sex.ch1 == "Male" & pheno.cn.ad$disease.state.ch1 %in% c("Alzheimer's disease" ,"control"),]
nb.samples.after.pca.male <- nrow(pheno.cn.ad.male)
```


```{R}
betaQN_BMIQ_PCfiltered <- readRDS(file.path(dir.data.processed, "addNeuroMed_se_withXY.rds"))
nb.samples.after.pca <- ncol(betaQN_BMIQ_PCfiltered)
```

```{R, echo = FALSE}
pheno <- colData(betaQN_BMIQ_PCfiltered) 
pheno <- pheno %>% as.data.frame %>% dplyr::select(
  c(disease.state.ch1, Sex.ch1, age.ch1, 
    B, NK,   CD4T, CD8T,   Mono, Neutro, Eosino
  )
)

pheno %>% str()

pheno %>%
  DT::datatable(
    filter = 'top',
    style = "bootstrap",
    extensions = 'Buttons',
    options = list(
      scrollX = TRUE, 
      dom = 'Bfrtip',
      buttons = I('colvis'),
      keys = TRUE, 
      pageLength = 10
    ), 
    rownames = FALSE,
    caption = "Samples metadata"
  )
```





## Numbers of samples and probes removed in each step

### Female
```{R}
df.samples <- data.frame(
  "Number of samples" =  c(
    nb.samples.age.65.exclusion.female, 
    nb.samples.after.pca.female
  ),
  "Description" = c(
    "Number of FEMALE samples before preprocessing",
    "Number of FEMALE samples after PCA"
  )
)    
df.samples 
```

### Male
```{R}
df.samples <- data.frame(
  "Number of samples" =  c(
    nb.samples.age.65.exclusion.male, 
    nb.samples.after.pca.male
  ),
  "Description" = c(
    "Number of MALE samples before preprocessing",
    "Number of MALE samples after PCA"
  )
)    
df.samples 
```


```{R}
# Create summary table
df.probes <- data.frame(
  "Number of probes" = c(
    nb.probes.before.norm,
    nb.probes.after.norm
  ),
  "Description" = c(
    "total number of probes in raw data",
    "only probes that start with cg, no crosshyb, include smoking probes, mafcut = 0.01"
  )
)
df.probes %>% gt::gt() 
```


# Checking the data

```{R checking_part1}
se <- readRDS(file.path(dir.data.processed, "addNeuroMed_se_withXY.rds"))
```


## show probes not starting with cg are removed
```{R checking_cg}
grep("cg",rownames(se), invert = T)
```

## show probes on X/Y are not removed
```{R checking_XY}
c("chrX", "chrY") %in% (seqnames(se) %>% as.character())
```

## show probes with SNP with MAF >= 0.01 and in the last 5bp were removed (compared with DMRcateData package probes)
```{R checking_maf}
eh <- ExperimentHub()
snpsall <- eh[["EH3130"]]
any(snpsall$probe[snpsall$mafs > 0.01 & snpsall$distances <= 5] %>% as.character() %in% rownames(se)) 
crosshyb <- eh[["EH3129"]]
any(crosshyb %in% rownames(se))
XY.probes <- eh[["EH3131"]]
any(XY.probes %in% rownames(se)) 
```

## select a few smoking probes, and show they are included
```{R checking_smoking}
any(
  c("cg16145216", "cg19406367",
    "cg05603985", "cg10768063",
    "cg14099685", "cg12513616",
    "cg04085571", "cg23648810"
  ) %in% rownames(se)
)
```

## show quantiles for each sample is the same
```{R checking_quantiles}
boxplot(assay(se[,1:10]))
plyr::adply(assay(se[,1:10]),.margins = 2,.fun = function(x){quantile(x)})
```


## check cell type proportions adds up to 1
```{R checking_cell_type_proportion}
table(rowSums(colData(se)[,c("B","NK","CD4T","CD8T","Mono","Neutro","Eosino")] %>% as.matrix()))
```

## check outliers samples are not in datset anymore
```{R checking_outliers}
outliers <- readr::read_csv(paste0(dir.data.pca, "addNeuroMed_PCs_usingBetas.csv"),col_types = readr::cols()) %>%
  dplyr::filter(outlier_PC1 == 1 | outlier_PC2 == 1) %>% pull(`...1`) 
head(outliers)
head(colnames(se))
any(outliers %in% colnames(se))
```

## for one sample, check distribution of type I and type II probes
```{R checking_part_probes_types, eval = TRUE}
library(sm)

annotType <- IlluminaHumanMethylation450kanno.ilmn12.hg19::Manifest
annotType$designTypeNumeric <- ifelse(annotType$Type == "I",1,2)
probes <- assay(se)[,1] %>% na.omit()
sm.density.compare(
  probes,
  annotType[names(probes),]$designTypeNumeric
)

probes <- assay(se)[,2] %>% na.omit()
sm.density.compare(
  probes,
  annotType[names(probes),]$designTypeNumeric
)
```

## check annotated sex is identical to predicted sex

```{r echo = F}
pheno <- data.frame(colData (se))
pheno$DIAGNOSIS <- factor(pheno$disease.state.ch1, levels = c("control", "mild cognitive impairment", "Alzheimer's disease"))
pheno$barcodes <- row.names(pheno)
pheno$Sex <- factor(pheno$Sex.ch1)
pheno <- pheno %>% dplyr::select(
  c(barcodes, Sex, age.ch1, 
    DIAGNOSIS, 
    B, NK,   CD4T, CD8T,   Mono, Neutro, Eosino
  )
)
```


```{r checking_sex_identical, message = FALSE}
library(wateRmelon)
betaSet <- assay(se)
est_Sex <- estimateSex(betaSet, do_plot = FALSE)
identical(factor(est_Sex$predicted_sex), pheno$Sex)
```


```{R checking_part5, include = FALSE}
rm(se)
```

# Analysis

```{R analysis_data}
addNeuroMed.se <- readRDS(file.path(dir.data.processed, "addNeuroMed_se_withXY.rds"))
result_SSc <- readr::read_csv(file.path(dir.data,"addNeuroMed_Smoking_Scores.csv"))

beta <- assay(addNeuroMed.se) * 100

#  0 = control, 1 = AD, 2 = MCI
pheno <- data.frame(colData (addNeuroMed.se))
pheno$DIAGNOSIS <- factor(pheno$disease.state.ch1, levels = c("control", "mild cognitive impairment", "Alzheimer's disease"))
pheno$barcodes <- row.names(pheno)
pheno$Sex <- factor(pheno$Sex.ch1)
pheno <- pheno %>% dplyr::select(
  c(barcodes, Sex, age.ch1, 
    DIAGNOSIS, 
    B, NK,   CD4T, CD8T,   Mono, Neutro, Eosino
  )
)
pheno$granulocytes <- pheno$Eosino + pheno$Neutro

pheno$smokingScore <- result_SSc[match(pheno$barcodes, result_SSc$SampleName),]$smokingScore
```



## Metadata structure
```{R}
str(pheno)
```

## Number of samples
```{R}
plyr::count(pheno$DIAGNOSIS)
```

```{R analysis_data_part2}

pheno.cn.ad <- pheno[pheno$DIAGNOSIS %in% c("control", "Alzheimer's disease"),]
pheno.cn.ad$DIAGNOSIS <-  forcats::fct_drop(pheno.cn.ad$DIAGNOSIS)
pheno.cn.ad.male <- pheno.cn.ad[pheno.cn.ad$Sex %in% "Male",]
pheno.cn.ad.female <- pheno.cn.ad[pheno.cn.ad$Sex %in% "Female",]

beta.cn.ad.male <- beta[, pheno.cn.ad.male$barcodes]
beta.cn.ad.female <- beta[, pheno.cn.ad.female$barcodes]

```

```{r}
plyr::count(pheno.cn.ad.female$DIAGNOSIS)
pheno.cn.ad.female %>% dplyr::group_by(DIAGNOSIS) %>% dplyr::summarise(mean(age.ch1), sd(age.ch1)) 
pheno.cn.ad.female %>% dplyr::summarise(mean(age.ch1), sd(age.ch1))
```

```{r}
plyr::count(pheno.cn.ad.male$DIAGNOSIS)
pheno.cn.ad.male %>% dplyr::group_by(DIAGNOSIS) %>% dplyr::summarise(mean(age.ch1), sd(age.ch1)) 
pheno.cn.ad.male %>% dplyr::summarise(mean(age.ch1), sd(age.ch1))
```

```{r}
x <- plyr::count(pheno.cn.ad,c("Sex","DIAGNOSIS"))
ggpubr::ggbarplot(data = x,x = "DIAGNOSIS",y = "freq",facet.by = "Sex",color = "DIAGNOSIS",fill = "DIAGNOSIS") + ggtitle("AddNeuroMed") +  
  theme(
    axis.title.x = element_blank(),                                                                                      axis.text.x = element_blank(),
    axis.ticks.x = element_blank())

```

## Wilcoxon test on smoking scores for AD vs controls


### Male

```{r}
SSc.male.ad <- pheno.cn.ad.male %>% filter(DIAGNOSIS == "Alzheimer's disease") %>% dplyr::select(smokingScore)
SSc.male.cn <- pheno.cn.ad.male %>% filter(DIAGNOSIS == "control") %>% dplyr::select(smokingScore)
wilcox.stats.male <- wilcox.test(SSc.male.ad$smokingScore, SSc.male.cn$smokingScore)
wilcox.stats.male
```

### Female

```{r}
SSc.female.ad <- pheno.cn.ad.female %>% filter(DIAGNOSIS == "Alzheimer's disease") %>% dplyr::select(smokingScore)
SSc.female.cn <- pheno.cn.ad.female %>% filter(DIAGNOSIS == "control") %>% dplyr::select(smokingScore)
wilcox.stats.female <- wilcox.test(SSc.female.ad$smokingScore, SSc.female.cn$smokingScore)
wilcox.stats.female
```

```{r plot_SSc, echo = F}
ggplot(pheno.cn.ad, aes(x = DIAGNOSIS,y = smokingScore, fill = DIAGNOSIS)) + 
  geom_boxplot() + geom_jitter(shape=16, position=position_jitter(0.2)) +
  ggpubr::stat_compare_means(method = "wilcox.test") +
  theme_classic() + theme(legend.position="top") + 
  scale_fill_brewer(palette="Dark2") +
  facet_grid(~Sex) + ylab('Smoking Scores')
```




## Logistic Regression Model -- Without adjusting smoking score


```{R analysis, eval = FALSE}
doParallel::registerDoParallel(cores = 10)

## Male

message("o Stating analysis - AD vs CN")
results.ad.glm.male <- plyr::adply(
  .data = beta.cn.ad.male, 
  .margins = 1,
  .fun = function(one){
    dat.cn.ad <- data.frame (beta = one, pheno.cn.ad.male)
    b <- glm(
      DIAGNOSIS ~ beta + age.ch1  + B + NK + CD4T + Mono + granulocytes,
      dat.cn.ad,
      family = binomial
    )
    summary(b)
    beta_results <- coef(summary(b)) [grep("beta",rownames(coef(summary(b)) ),value = TRUE),]
    return(beta_results)
  },.parallel = TRUE, .progress = "time",.id = "cpg",.inform = FALSE
)

readr::write_csv(
  x = results.ad.glm.male ,
  file = file.path(dir.results.glm, "MALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results.csv")
)

## Female

results.ad.glm.female <- plyr::adply(
  .data = beta.cn.ad.female, 
  .margins = 1,
  .fun = function(one){
    dat.cn.ad <- data.frame (beta = one, pheno.cn.ad.female)
    b <- glm(
      DIAGNOSIS ~ beta + age.ch1  + B + NK + CD4T + Mono + granulocytes,
      dat.cn.ad,
      family = binomial
    )
    summary(b)
    beta_results <- coef(summary(b)) [grep("beta",rownames(coef(summary(b)) ),value = TRUE),]
    return(beta_results)
  },.parallel = TRUE, .progress = "time",.id = "cpg",.inform = FALSE
)

readr::write_csv(
  x = results.ad.glm.female ,
  file = file.path(dir.results.glm, "FEMALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results.csv")
)
```

## Logistic Regression Model -- Adjust smoking score


```{R analysis2, eval = FALSE}
doParallel::registerDoParallel(cores = 10)

## Male

message("o Stating analysis - AD vs CN")
results.ad.glm.male.SSc <- plyr::adply(
  .data = beta.cn.ad.male, 
  .margins = 1,
  .fun = function(one){
    dat.cn.ad <- data.frame (beta = one, pheno.cn.ad.male)
    b <- glm(
      DIAGNOSIS ~ beta + age.ch1  + B + NK + CD4T + Mono + granulocytes + smokingScore,
      dat.cn.ad,
      family = binomial
    )
    summary(b)
    beta_results <- coef(summary(b)) [grep("beta",rownames(coef(summary(b)) ),value = TRUE),]
    return(beta_results)
  },.parallel = TRUE, .progress = "time",.id = "cpg",.inform = FALSE
)

readr::write_csv(
  x = results.ad.glm.male.SSc ,
  file = file.path(dir.results.glm, "MALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_results.csv")
)

## Female

message("o Stating analysis - AD vs CN")
results.ad.glm.female <- plyr::adply(
  .data = beta.cn.ad.female, 
  .margins = 1,
  .fun = function(one){
    dat.cn.ad <- data.frame (beta = one, pheno.cn.ad.female)
    b <- glm(
      DIAGNOSIS ~ beta + age.ch1  + B + NK + CD4T + Mono + granulocytes + smokingScore,
      dat.cn.ad,
      family = binomial
    )
    summary(b)
    beta_results <- coef(summary(b)) [grep("beta",rownames(coef(summary(b)) ),value = TRUE),]
    return(beta_results)
  },.parallel = TRUE, .progress = "time",.id = "cpg",.inform = FALSE
)

readr::write_csv(
  x = results.ad.glm.female ,
  file = file.path(path.to.hpc, path.result.glm, "FEMALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_results.csv")
)
```



## Annotate results
```{R annotate, eval = FALSE}
message("o Stating to annotate")
addNeuroMed.se <- readRDS(file.path(dir.data.processed, "addNeuroMed_se_withXY.rds"))

load(file.path(dir.data.aux,"great_EPIC_array_annotation.rda"))
load(file.path(dir.data.aux,"E073_15_coreMarks_segments.rda"))
load(file.path(dir.data.aux,"meta_analysis_cpgs.rda"))

annotate_results <- function(result, data.se){
  result$fdr <- p.adjust(result$`Pr(>|z|)`,method = "fdr")
  result <- cbind(
    result,
    rowRanges(data.se)[result$cpg] %>% as.data.frame() %>% dplyr::select(1:4)
  )
  result$Islands.UCSC.Relation_to_Island <- IlluminaHumanMethylation450kanno.ilmn12.hg19::Islands.UCSC[result$cpg,"Relation_to_Island"]
  result$UCSC_RefGene_Name <- IlluminaHumanMethylation450kanno.ilmn12.hg19::Other[result$cpg,"UCSC_RefGene_Name"]       
  result$UCSC_RefGene_Group <- IlluminaHumanMethylation450kanno.ilmn12.hg19::Other[result$cpg,"UCSC_RefGene_Group"]     
  
  result <- dplyr::left_join(result, great,by = c("seqnames","start","end","cpg"))
  
  hits <- findOverlaps(rowRanges(data.se)[result$cpg],ChmmModels.gr) %>% as.data.frame()
  hits$state <- ChmmModels.gr$state[hits$subjectHits]
  hits$cpg <- result$cpg[hits$queryHits]
  result <- merge(result, hits[,c("state","cpg")], all.x = TRUE,sort = FALSE, by = "cpg")
  
  # sig.in.brain: 1 if it overlaps with the 3751 CpGs or 119 DMRs from our brain samples meta-analysis
  result$sig.in.brain <- result$cpg %in% brain.meta.analysis.cpgs
  return(result)
}

colnames(results.ad.glm.male)[1] <- "cpg"
results.ad.glm.male.annotated <- results.ad.glm.male %>% annotate_results(addNeuroMed.se)
readr::write_csv(
  x = results.ad.glm.male.annotated,
  file = file.path(dir.results.glm, "MALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)

colnames(results.ad.glm.female)[1] <- "cpg"
results.ad.glm.annotated.female <- results.ad.glm.female %>% annotate_results(addNeuroMed.se)
readr::write_csv(
  x = results.ad.glm.annotated.female,
  file = file.path(dir.results.glm,"FEMALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)

colnames(results.ad.glm.male.SSc)[1] <- "cpg"
results.ad.glm.male.annotated.SSc <- results.ad.glm.male.SSc %>% annotate_results(addNeuroMed.se)
readr::write_csv(
  x = results.ad.glm.male.annotated.SSc,
  file = file.path(dir.results.glm, "MALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)

colnames(results.ad.glm.female.SSc)[1] <- "cpg"
results.ad.glm.annotated.female.SSc <- results.ad.glm.female.SSc %>% annotate_results(addNeuroMed.se)
readr::write_csv(
  x = results.ad.glm.annotated.female.SSc,
  file = file.path(dir.results.glm,"FEMALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)
```


```{R read_results, include = FALSE}
results.ad.glm.male <- readr::read_csv(
  file.path(dir.results.glm,"MALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)
results.ad.glm.female <- readr::read_csv(
  file.path(dir.results.glm,"FEMALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)
results.ad.glm.male.SSc <- readr::read_csv(
  file.path(dir.results.glm,"MALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)
results.ad.glm.female.SSc <- readr::read_csv(
  file.path(dir.results.glm,"FEMALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated.csv")
)
```

# Calculate inflation factors and Bacon correction for final results

```{R, message = FALSE, warning = FALSE, result = "hide"}
library(dplyr)
library(bacon)
library(GWASTools)
```

## Auxiliary function

```{R inflation}
estimation_of_inflation <- function(data){
  ### 1. Compute genomic inflation factor before bacon adjustment
  data$zvalue <- data$Estimate / data$`Std. Error`
  #data$zvalue <- qnorm(pt(data$tvalue, df))
  data$chisq <- (data$zvalue) ^ 2
  # inflation factor - last term is median from chisq distrn with 1 df  
  inflationFactor <- median(data$chisq,na.rm = TRUE) / qchisq(0.5, 1)
  print("lambda")
  print(inflationFactor)
  # genome-wide sig cpgs
  #sig <- ifelse(data$`Pr(>|t|)` < 2.4e-7, 1, 0)
  # table(sig)  # 1 sig
  ### 2. bacon analysis
  bc <- bacon(
    teststatistics = NULL,
    effectsizes =  data$Estimate,
    standarderrors = data$`Std. Error`,
    na.exclude = TRUE
  )
  # inflation factor
  print("lambda.bacon")
  print(inflation(bc))
  # bias
  print("estimated bias")
  print(bias(bc))
  
  ### 3. Create final dataset
  data.with.inflation <- data.frame(
    data,
    Estimate.bacon = bacon::es(bc),
    StdErr.bacon = bacon::se(bc),
    pValue.bacon = pval(bc),
    fdr.bacon = p.adjust(pval(bc), method = "fdr"),
    stringsAsFactors = FALSE
  )
  data.with.inflation <- data.with.inflation %>% dplyr::select(-c(zvalue, chisq))
  return(
    list(
      "data.with.inflation" = data.with.inflation,
      "inflationFactor" =   inflationFactor,
      "estimatedInflation" = inflation(bc),
      "estimatedBias" = bias(bc)
    )
  )
}
```

## No adjust smoking

```{R estimation_of_inflation_ad1}
results.ad.glm.male.with.inflation <- estimation_of_inflation(data = results.ad.glm.male)
```

```{r eval = F}
readr::write_csv(
  x = results.ad.glm.male.with.inflation$data.with.inflation,
  file = file.path(dir.results.glm,"MALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_annotated_with_inflation.csv")
)
```


```{r}
qqPlot(
  results.ad.glm.male.with.inflation$data.with.inflation$Pr...z.., ci = TRUE, 
  main = paste0(cohort, " cohort")
) + text(
  x = 0.5, 
  y = 3,
  labels = bquote(lambda == .(format(results.ad.glm.male.with.inflation$inflationFactor,digits = 4))),
  pos = 4
) + text(
  x = 0.5, 
  y = 2.5,
  labels = bquote(lambda[bacon] == .(format(results.ad.glm.male.with.inflation$estimatedInflation,digits = 4))),
  pos = 4
)
```

```{R estimation_of_inflation_ad2}
results.ad.glm.female.with.inflation <- estimation_of_inflation(data = results.ad.glm.female)
```

```{r eval = F}
readr::write_csv(
  x = results.ad.glm.female.with.inflation$data.with.inflation,
  file = file.path(dir.results.glm,"FEMALE_addNeuroMed_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_annotated_with_inflation.csv")
)
```

```{r}
qqPlot(
  results.ad.glm.female.with.inflation$data.with.inflation$Pr...z.., ci = TRUE, 
  main = paste0(cohort, " cohort")
) + text(
  x = 0.5, 
  y = 3,
  labels = bquote(lambda == .(format(results.ad.glm.female.with.inflation$inflationFactor,digits = 4))),
  pos = 4
) + text(
  x = 0.5, 
  y = 2.5,
  labels = bquote(lambda[bacon] == .(format(results.ad.glm.female.with.inflation$estimatedInflation,digits = 4))),
  pos = 4
)
```


## Adjust smoking

```{R estimation_of_inflation_ad3}
results.ad.glm.male.with.inflation.SSc <- estimation_of_inflation(data = results.ad.glm.male.SSc)
```

```{r eval = F}
readr::write_csv(
  x = results.ad.glm.male.with.inflation.SSc$data.with.inflation,
  file = file.path(dir.results.glm,"MALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_annotated_with_inflation.csv")
)
```

```{r}
qqPlot(
  results.ad.glm.male.with.inflation.SSc$data.with.inflation$Pr...z.., ci = TRUE, 
  main = paste0(cohort, " cohort")
) + text(
  x = 0.5, 
  y = 3,
  labels = bquote(lambda == .(format(results.ad.glm.male.with.inflation.SSc$inflationFactor,digits = 4))),
  pos = 4
) + text(
  x = 0.5, 
  y = 2.5,
  labels = bquote(lambda[bacon] == .(format(results.ad.glm.male.with.inflation.SSc$estimatedInflation,digits = 4))),
  pos = 4
)
```

```{R estimation_of_inflation_ad4}
results.ad.glm.female.with.inflation.SSc <- estimation_of_inflation(data = results.ad.glm.female.SSc)
```

```{r eval = F}
readr::write_csv(
  x = results.ad.glm.female.with.inflation.SSc$data.with.inflation,
  file = file.path(dir.results.glm,"FEMALE_addNeuroMed_glm_diagnosis_vs_beta_smoking_and_covaritates_age_at_least_65_AD_vs_CN_annotated_with_inflation.csv")
)
```

```{r}

qqPlot(
  results.ad.glm.female.with.inflation.SSc$data.with.inflation$Pr...z.., ci = TRUE, 
  main = paste0(cohort, " cohort")
) + text(
  x = 0.5, 
  y = 3,
  labels = bquote(lambda == .(format(results.ad.glm.female.with.inflation.SSc$inflationFactor,digits = 4))),
  pos = 4
) + text(
  x = 0.5, 
  y = 2.5,
  labels = bquote(lambda[bacon] == .(format(results.ad.glm.female.with.inflation.SSc$estimatedInflation,digits = 4))),
  pos = 4
)
```


# Smoking effects 

## spearman correlation


```{r}
effect.size.ad.glm.bacon <- data.frame(
  non.Adjust.Smoke.male = results.ad.glm.male.with.inflation$data.with.inflation$Estimate.bacon,
  Adjust.Smoke.male = results.ad.glm.male.with.inflation.SSc$data.with.inflation$Estimate.bacon,
  non.Adjust.Smoke.female = results.ad.glm.female.with.inflation$data.with.inflation$Estimate.bacon,
  Adjust.Smoke.female = results.ad.glm.female.with.inflation.SSc$data.with.inflation$Estimate.bacon
  )
```


```{r}
## male 
cor.test(effect.size.ad.glm.bacon$non.Adjust.Smoke.male, effect.size.ad.glm.bacon$Adjust.Smoke.male,
         method = 'spearman')
```

```{r}
## female 
cor.test(effect.size.ad.glm.bacon$non.Adjust.Smoke.female, effect.size.ad.glm.bacon$Adjust.Smoke.female,
         method = 'spearman')
```

### Scatter plot

```{r echo = F, message = F}
library(ggpubr)
## male 
ggscatter(data = effect.size.ad.glm.bacon, x = 'non.Adjust.Smoke.male',
          y = 'Adjust.Smoke.male', size = 0.5, cor.coef	= T, cor.method = "spearman",
          xlab = 'Effect size of non-adjust smoking',
          ylab = 'Effect size of adjusted smoking',
          title = 'Male') 
```
```{r echo = F, message = F}
## Female
ggscatter(data = effect.size.ad.glm.bacon, x = 'non.Adjust.Smoke.female',
          y = 'Adjust.Smoke.female', size = 0.5, cor.coef	= T, cor.method = "spearman",
          xlab = 'Effect size of non-adjust smoking',
          ylab = 'Effect size of adjusted smoking',
          title = 'Female') 
```



## Venn diagram 

### Significant cpgs (with bacon pvalue <1E-3) 


```{R venn_bacon, fig.width = 8, fig.height = 8, message = F}
library(VennDiagram)

list <-  list(
  "\nMALE AD vs CN \nwithout adjust smoking\n(Bacon p-values < 1E-3)" = results.ad.glm.male.with.inflation$data.with.inflation$cpg[results.ad.glm.male.with.inflation$data.with.inflation$pValue.bacon < 1E-3],
  "\nMALE AD vs CN \nadjusted smoking\n(Bacon p-values < 1E-3)" = results.ad.glm.male.with.inflation.SSc$data.with.inflation$cpg[results.ad.glm.male.with.inflation.SSc$data.with.inflation$pValue.bacon < 1E-3]
)
venn <- venn.diagram(list, filename = NULL, 
                     col = "transparent",
                     fill = c("blue", "green"),
                     alpha = 0.30, scaled = F, 
                     direct.area = F, 
                     sub.cex = 0.3,
                     height = 480 , 
                     width = 480 , 
                     cat.pos = c(-27, 27),
                     print.mode=c("raw","percent"),
                     output=TRUE,compression = "lzw")
grid.draw(venn)
```

```{R venn_bacon2, fig.width = 8, fig.height = 8}
list <-  list(
  "\nFEMALE AD vs CN \nwithout adjust smoking\n(Bacon p-values < 1E-3)" = results.ad.glm.female.with.inflation$data.with.inflation$cpg[results.ad.glm.female.with.inflation$data.with.inflation$pValue.bacon < 1E-3],
  "\nFEMALE AD vs CN \nadjusted smoking\n(Bacon p-values < 1E-3)" = results.ad.glm.female.with.inflation.SSc$data.with.inflation$cpg[results.ad.glm.female.with.inflation.SSc$data.with.inflation$pValue.bacon < 1E-3]
)
venn <- venn.diagram(list, filename = NULL, 
                     col = "transparent",
                     fill = c("blue", "green"),
                     alpha = 0.30, scaled = F, 
                     direct.area = F, 
                     sub.cex = 0.3,
                     height = 480 , 
                     width = 480 , 
                     print.mode=c("raw","percent"),
                     output=TRUE,compression = "lzw")
grid.draw(venn)
```

# Session information
```{R}
devtools::session_info()
```