---
title: "Distinct sex-specific DNA methylation differences in Alzheimer’s disease"
subtitle: 'ADNI dataset'
author:
  - Tiago Chedraoui Silva^[University of Miami]
  - Wei Zhang^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
knitr::opts_knit$set(root.dir = "./../../")
```

# Setup

```{R Setup, message = FALSE, results = 'hide'}
library(dplyr)
library(SummarizedExperiment)
library(ExperimentHub)
library(readxl)
library(wateRmelon)
library(minfi)
library(ggpubr)
library(ggrepel)
cohort <- "ADNI"
dir.base <- getwd()
dir.base.data <- file.path(dir.base,"../AD-meta-analysis-blood-samples/")
dir.base.data2 <- file.path(dir.base,"../AD-meta-analysis-blood-samples-bySex/")
dir.data <- file.path(dir.base.data,"datasets/",cohort,"/data/DNA_methylation") 
dir.data.clinical <- file.path(dir.base.data,"datasets/",cohort,"/data/Clinical") 
dir.data.raw <- file.path(dir.base.data,"datasets/",cohort,"/data/DNA_methylation/idat") 
dir.data.processed0 <- file.path(dir.base.data,"datasets/",cohort,"/data/DNA_methylation/processed") 
dir.data.processed <- file.path(dir.base.data,"datasets/",cohort,"/data/DNA_methylation/processed/withSmoke/") 
dir.data.aux <- file.path(dir.base.data,"datasets/Aux/") 
dir.data.pca <- file.path(dir.data,"/pca_filtering/withSmoke/") 
dir.results <- file.path(dir.base.data2,"analysis_results",cohort,"/withSmoke")
#dir.results.glm <- file.path(dir.results,"/Logistic_regression_model")
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

# Data retrieval

```{R, load_rgset_bs_filtered_min_65, include = T, eval = FALSE}
load(file.path(dir.data.processed0,"RGSet_bs_filtered_min_age_at_visit_65_AD_CN.rda"))
```

```{R, load_bs_filtered_min_65, include = T, eval = FALSE}
adni.se <- readRDS(file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65_AD_CN.RDS"))
```


```{R, load_detectP, include = FALSE, eval = FALSE}
load(file.path(dir.data.processed0,"detectionP_passed_probes_AD_CN.rda"))
nb.probes.before.norm <- nrow(failed.01)
# length(passedProbes) #833642
# length(grep("cg",passedProbes)) #831149
```

## Probes QC

1. keep only probes that start with "cg"
2. drop probes where SNP with MAF >= 0.01 was present in the last 5 bp of the probe.

```{R, eval = FALSE}
library(minfi)
library(wateRmelon)
library(RPMM)
library(DMRcate)
```

```{R, eval = FALSE}
annotation <- minfi::getAnnotation(RGSet.bs.filtered.min.age_at_visit.65.AD_CN) %>% as.data.frame %>% dplyr::filter(chr %in% c("chrX","chrY"))
betaSet <- getBeta(RGSet.bs.filtered.min.age_at_visit.65.AD_CN)
betaSet <- betaSet[rownames(betaSet) %in% rownames(annotation), ]
nb.chrXY.probes <- nrow(betaSet)
rm(RGSet.bs.filtered.min.age_at_visit.65.AD_CN)
betaSet <- betaSet[rownames(betaSet) %in% passedProbes, ]
betaSet <- betaSet[grep("cg",rownames(betaSet)),]
gc()

# snapshotDate(): 2021-03-30
betaSet <- rmSNPandCH(
  object = betaSet,
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = TRUE,
  rmXY = FALSE
)
```

```{R, include = F}
# > dim(betaSet)
# [1] 16200  822
```


## Merge DNA methylation matrix on autosome with the one on chrX, chrY

```{R, eval = FALSE}
betaSet <- betaSet[,colnames(adni.se)]
identical(colnames(adni.se), colnames(betaSet))
betaQN_BMIQ <- rbind(assay(adni.se), betaSet)
table(base::duplicated(row.names(betaQN_BMIQ)))
# 766978
save(
  betaQN_BMIQ,
  clinical.filtered.min.age_at_visit.65.AD_CN,
  file = file.path(dir.data.processed,"RGSet_bs_filtered_min_age_at_visit_65_with_XY_AD_CN.rda")
)
```

### Summarize Experiment

```{R, eval = FALSE}
#library(ADNIMERGE)
library(dplyr)
library(readxl)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(S4Vectors)
epic.hg19 <- minfi::getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
epic.hg19.gr <- epic.hg19 %>% makeGRangesFromDataFrame(
  start.field = "pos", end.field = "pos",keep.extra.columns = T
)
clinical <- colData(adni.se)
betaQN_BMIQ <- betaQN_BMIQ[rownames(betaQN_BMIQ) %in% names(epic.hg19.gr),]
rowData <- epic.hg19.gr[rownames(betaQN_BMIQ),]
data <- betaQN_BMIQ[,colnames(betaQN_BMIQ) %in% rownames(clinical)]
data <- data[,rownames(clinical)]

adni.se <- SummarizedExperiment::SummarizedExperiment(
  assays = SimpleList("DNAm" = data),
  rowData = rowData,
  metadata = list("genome" = "hg19"),
  colData = clinical
)

saveRDS(adni.se, file = file.path(dir.data.processed,"adni_se_min_age_at_visit_65_with_XY_AD_CN.RDS"))
readr::write_csv(
  clinical %>% as.data.frame, 
  file = file.path(dir.data.processed,"ADNI_Se_colData_with_XY_AD_CN.csv")
)
```


# Summary after QC steps


## Data and metadata
```{R}
adni.se <- readRDS(file.path(dir.data.processed,"adni_se_min_age_at_visit_65_with_XY_AD_CN.RDS"))
#adni.se <- adni.se[,adni.se$DX %in% c("CN","Dementia")]
nb.samples.after.pca <- ncol(adni.se) #793
nb.probes.after.norm <- nrow(adni.se) #766978
```

```{R}
adni.prediction <- wateRmelon::estimateSex(assay(adni.se))
plyr::count(adni.prediction$predicted_sex != adni.se$PTGENDER)
colData(adni.se[,adni.prediction$predicted_sex != adni.se$PTGENDER])
adni.prediction[adni.prediction$predicted_sex != adni.se$PTGENDER,]
```



```{R}
pheno.cn.ad <- colData(adni.se)
pheno.cn.ad.female <- pheno.cn.ad[pheno.cn.ad$PTGENDER == "Female" & pheno.cn.ad$DX %in% c("CN","Dementia"),]
nb.samples.after.pca.female <- nrow(pheno.cn.ad.female) #364
plyr::count(pheno.cn.ad.female$DX)
plyr::count(pheno.cn.ad.female %>% as.data.frame() %>% group_by(RID) %>% dplyr::slice(which.max(DateDrawn)) %>% pull(DX)) ## count last visit ## 179
pheno.cn.ad.female %>% as.data.frame %>% dplyr::group_by(DX) %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit))
pheno.cn.ad.female %>% as.data.frame %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit))  %>% round(digits = 1)
```

```{R}
pheno.cn.ad.male <- pheno.cn.ad[pheno.cn.ad$PTGENDER == "Male" & pheno.cn.ad$DX %in% c("CN","Dementia"),]
plyr::count(pheno.cn.ad.male$DX)
nb.samples.after.pca.male <- nrow(pheno.cn.ad.male) #429
plyr::count(pheno.cn.ad.male %>% as.data.frame() %>% group_by(RID) %>% dplyr::slice(which.max(DateDrawn)) %>% pull(DX)) ## count last visit ## 219
pheno.cn.ad.male %>% as.data.frame %>% dplyr::group_by(DX) %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit))
pheno.cn.ad.male %>% as.data.frame %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit))  %>% round(digits = 1)
```


## Numbers of samples and probes removed in each step

### Female
```{R, eval = FALSE, echo=FALSE, include = F}
df.samples <- data.frame(
  "Number of samples" =  c(
    nb.samples, 
    nb.samples.bc.filtered, 
    nb.samples.before.pca, 
    nb.samples.after.pca.female
  ),
  "Description" = c(
    "Number of FEMALE samples before bisulfiteConversion filter",
    "Number of FEMALE samples after bisulfiteConversion filter",
    "Number of FEMALE samples before PCA",
    "Number of FEMALE samples after PCA"
  )
)    
df.samples 
```

### Male
```{R, eval = FALSE, echo=FALSE, include = F}
df.samples <- data.frame(
  "Number of samples" =  c(
    nb.samples, 
    nb.samples.bc.filtered, 
    nb.samples.before.pca, 
    nb.samples.after.pca.male
  ),
  "Description" = c(
    "Number of MALE samples before bisulfiteConversion filter",
    "Number of MALE samples after bisulfiteConversion filter",
    "Number of MALE samples before PCA",
    "Number of MALE samples after PCA"
  )
)    
df.samples 
```


```{R, eval = FALSE, echo=FALSE, include = F}
# Create summary table
df.probes <- data.frame(
  "Number of probes" = c(
    nb.probes.before.norm,
    length(passedProbes),
    nb.probes.after.norm
  ),
  "Description" = c(
    "total number of probes in raw data",
    "probes with detection P < 0.01 for all samples",
    "only probes that start with cg, no crosshyb, mafcut = 0.01"
  )
)
df.probes %>% gt::gt() 
```

# Checking the data

## show probes not starting with cg are removed
```{R check_cg}
grep("cg",rownames(adni.se),invert = T)
```

## show probes on X/Y are included
```{R check_xy}
c("chrX", "chrY") %in% (seqnames(adni.se) %>% as.character())
```

## show probes with SNP with MAF >= 0.01 and in the last 5bp were removed (compared with DMRcateData package probes) 
```{R}
eh <- ExperimentHub()
snpsall <- eh[["EH3130"]]
any(snpsall$probe[snpsall$mafs > 0.01 & snpsall$distances <= 5] %>% as.character() %in% rownames(adni.se)) 
crosshyb <- eh[["EH3129"]]
any(crosshyb %in% rownames(adni.se))
XY.probes <- eh[["EH3131"]]
any(XY.probes %in% rownames(adni.se)) 
```

## select a few smoking probes, and show they are included
```{R}
any(
  c("cg16145216", "cg19406367",
    "cg05603985", "cg10768063",
    "cg14099685", "cg12513616",
    "cg04085571", "cg23648810"
  ) %in% rownames(adni.se)
)
```

## show quantiles for each sample is the same
```{R quantile_box_plot}
boxplot(assay(adni.se[,1:10]))
plyr::adply(assay(adni.se[,1:10]),.margins = 2,.fun = function(x){quantile(x)})
```


## check cell type proportions adds up to 1
```{R}
table(rowSums(colData(adni.se)[,c("B","NK","CD4T","CD8T","Mono","Neutro","Eosino")] %>% as.matrix()))
```

## show plate is a factor
```{R}
class(adni.se$PlateNumber)
```

## check outliers samples are not in datset anymore
```{R}
outliers <- readr::read_csv(paste0(dir.data.pca,
                                   "ADNI_PCs_usingBetas_min_age_at_visit_65_AD_CN.csv"),
                            col_types = readr::cols()) %>%
  dplyr::filter(outlier_PC1 == 1 | outlier_PC2 == 1) %>% pull(1) 
head(outliers)
head(colnames(adni.se))
any(outliers %in% colnames(adni.se))
```

## For one sample, check distribution of type I and type II probes
```{R checking_part2, eval = TRUE}
library(sm)

annotType <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Manifest
annotType$designTypeNumeric <- ifelse(annotType$Type == "I",1,2)
probes <- assay(adni.se)[,1] %>% na.omit()
sm.density.compare(
  probes,
  annotType[names(probes),]$designTypeNumeric
)

probes <- assay(adni.se)[,2] %>% na.omit()
sm.density.compare(
  probes,
  annotType[names(probes),]$designTypeNumeric
)

```


```{R checking_part5, include = FALSE}
rm(adni.se)
```


# Analysis
```{R load_analysis_data, eval = FALSE}
adni.se <- readRDS(file.path(dir.data.processed,"adni_se_min_age_at_visit_65_with_XY_AD_CN.RDS"))
```

```{r}
pheno <- data.frame(colData (adni.se))

str(pheno)

pheno$barcodes <- row.names(pheno)
pheno$Sex <- factor(pheno$PTGENDER)
pheno$PLATE <- factor(pheno$PlateNumber)
pheno$granulocytes <- pheno$Eosino + pheno$Neutro
pheno <- pheno %>% dplyr::select(
  c(RID, barcodes, Sex, age_at_visit, 
    DX, PlateNumber, PTEDUCAT,
    B, NK,   CD4T,   Mono, granulocytes
  )
)

pheno.cn.ad.male <- pheno[ pheno$Sex == "Male",]
pheno.cn.ad.female <- pheno[ pheno$Sex == "Female",]
```

```{R}
plyr::count(pheno.cn.ad.female$DX)

pheno.cn.ad.female.last_visit <- pheno.cn.ad.female %>% as.data.frame %>%  dplyr::group_by(RID) %>% filter(age_at_visit == max(age_at_visit))
plyr::count(pheno.cn.ad.female.last_visit$DX)
pheno.cn.ad.female.last_visit %>% dplyr::group_by(DX) %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit)) 
pheno.cn.ad.female.last_visit[,-1] %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit)) 
pheno.cn.ad.female.last_visit[,-1] %>% dplyr::group_by(DX) %>% summarise(mean(PTEDUCAT), sd(PTEDUCAT))
```



```{r}
## Female
pheno.cn.ad.female.last_visit  %>%
  t.test(
    PTEDUCAT ~ DX,
    data = .
  )
```

```{r}
## Male
pheno.cn.ad.male.last_visit  %>%
  t.test(
    PTEDUCAT ~ DX,
    data = .
  )
```


```{R}
plyr::count(pheno.cn.ad.male$DX)

pheno.cn.ad.male.last_visit <- pheno.cn.ad.male %>% as.data.frame %>%  dplyr::group_by(RID) %>% filter(age_at_visit == max(age_at_visit))
plyr::count(pheno.cn.ad.male.last_visit$DX)
pheno.cn.ad.male.last_visit %>% dplyr::group_by(DX) %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit)) 
pheno.cn.ad.male.last_visit[,-1] %>% dplyr::summarise(mean(age_at_visit), sd(age_at_visit)) 
pheno.cn.ad.male.last_visit[,-1] %>% dplyr::group_by(DX) %>% summarise(mean(PTEDUCAT), sd(PTEDUCAT))
```

```{r}
pheno.cn.ad.female.first_visit <- pheno.cn.ad.female %>% as.data.frame %>%  dplyr::group_by(RID) %>% filter(age_at_visit == min(age_at_visit))
plyr::count(pheno.cn.ad.female.first_visit$DX)
```


```{r, eval = T}
result_SSc <- readr::read_csv(file.path(dir.data,"ADNI_Smoking_Scores_AD_CN.csv"))
any(is.na(result_SSc$smokingScore))
```


```{R sas, eval = FALSE}
str_data <- function(se){
  
  mvalues <- log2(assay(se)/(1 - assay(se)))
  mvalues <- mvalues %>% t %>% as.data.frame()
  mvalues$barcodes <- rownames(mvalues)
  mvalues.pivot <- tidyr::pivot_longer(mvalues,!barcodes,names_to = "cpg", values_to = "mvalues")
  
  beta <- assay(se)
  beta <- beta %>% t %>% as.data.frame()
  beta$barcodes <- rownames(beta)
  beta.pivot <- pivot_longer(beta,!barcodes,names_to = "cpg", values_to = "beta")
  
  
  #  0 = control, 1 = AD, 2 = MCI
  pheno <- data.frame(colData (se))
  pheno$DIAGNOSIS <- factor(pheno$DX, levels = c("CN", "Dementia"))
  pheno <- pheno[!is.na(pheno$DIAGNOSIS),]
  pheno$barcodes <- row.names(pheno)
  
  # add indicator for first and last visit
  pheno$visit <- sub("m",  "", pheno$VISCODE)
  pheno$PlateNumber <- factor(pheno$PlateNumber)
  pheno$granulocytes <- pheno$Eosino + pheno$Neutro
  pheno$smokingScore <- result_SSc[match(pheno$barcodes, result_SSc$SampleName),]$smokingScore
  
  pheno <- pheno %>% dplyr::select(
    c(barcodes, RID, VISCODE, PTGENDER, age_at_visit, visit, 
      DIAGNOSIS, PlateNumber,
      B, NK,   CD4T, Mono,   granulocytes, smokingScore
    )
  )
  final <- dplyr::inner_join(pheno,beta.pivot) %>%  dplyr::inner_join(mvalues.pivot)
  final <- final[order(final$cpg,final$RID,final$VISCODE),]
  final
}

dir.create(file.path(dir.results,"/AD_CN//AD_CN_male/DATA_for_SAS_models//"),recursive = T)
for(start in seq(1,nrow(adni.se),by = 1000)){
  print(start)
  end <- start + 1000 - 1
  idx <- which(start == seq(1,nrow(adni.se),by = 1000))
  if(end > nrow(adni.se)) end <- nrow(adni.se)
  adni.se.selected <- adni.se[start:end,  adni.se$PTGENDER == "Male"]
  print(dim(adni.se.selected))
  final <- str_data(adni.se.selected)
  final$VISCODE <- final$visit <- final$mvalues <- NULL
  readr::write_csv(
    final,
    file = paste0(
      dir.results,
      "/AD_CN//AD_CN_male/DATA_for_SAS_models//",
      "ADNI_1000_cpgs_with_smok_probes",idx,".csv"
    )
  )
  #slackr::slackr_msg(paste("SAS data for male", idx, "saving completed!"))
}

dir.create(file.path(dir.results,"//AD_CN//AD_CN_female/DATA_for_SAS_models//"),recursive = T)
plyr::a_ply(seq(1,nrow(adni.se),by = 1000),.margins = 1,.fun = function(start){
  end <- start + 1000 - 1
  idx <- which(start == seq(1,nrow(adni.se),by = 1000))
  if(end > nrow(adni.se)) end <- nrow(adni.se)
  adni.se.selected <- adni.se[start:end, adni.se$PTGENDER == "Female"]
  final <- str_data(adni.se.selected)
  final$VISCODE <- final$visit <- final$mvalues <- NULL
  readr::write_csv(
    final,
    file = paste0(
      dir.results,
      "/AD_CN//AD_CN_female/DATA_for_SAS_models//",
      "ADNI_1000_cpgs_with_smok_probes",idx,".csv"
    )
  )
   #slackr::slackr_msg(paste("SAS data for female", idx, "saving completed!"))
},.progress = "time")
```



## Logistic regression model


### No adjust for smoking scores

```{r}
dir.results.glmm <- file.path(dir.results,"/AD_CN/Logistic_regression_model/")
```


#### Male
```{R, eval = FALSE}
dir.create(dir.results.glmm, recursive = T)
glmm_male_result <- read_xlsx(
  file.path(dir.base.data2,"LW/Males_SAS/results/ad-vs-cn-males_6-25-2022.xlsx"))
logistic_mixed_model_ad_cn.males <- glmm_male_result
logistic_mixed_model_ad_cn.males$file_order <- NULL
readr::write_csv(
  logistic_mixed_model_ad_cn.males, 
  file = file.path(dir.results.glmm, "Males_ADNI_AD_vs_CN_glmm_using_beta_SAS.csv")
)
```

#### Female
```{R, eval = FALSE}
glmm_female_result <- read_xlsx(
  file.path(dir.base.data2,"LW/Females_SAS/results/ad-vs-cn-females_6-25-2022.xlsx"))
logistic_mixed_model_ad_cn.females <- glmm_female_result
logistic_mixed_model_ad_cn.females$file_order <- NULL
readr::write_csv(
  logistic_mixed_model_ad_cn.females, 
  file = file.path(dir.results.glmm, "Females_ADNI_AD_vs_CN_glmm_using_beta_SAS.csv")
)
```

### Adjust for smoking scores

#### Males

```{r eval = F}
glmm_male_result <- read_xlsx(
  file.path(dir.base.data2,"LW/Males_SAS_adjust_smoking/results/ad-vs-cn-males-adjust-smoking_6-28-2022.xlsx"))
logistic_mixed_model_ad_cn.males.SSc <- glmm_male_result
logistic_mixed_model_ad_cn.males.SSc$file_order <- NULL
readr::write_csv(
  logistic_mixed_model_ad_cn.males.SSc, 
  file = file.path(dir.results.glmm, "Males_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_smoking.csv")
)
```

#### Females

```{r eval = F}
glmm_female_result <- read_xlsx(
  file.path(dir.base.data2,"LW/Females_SAS_adjust_smoking/results/ad-vs-cn-females-adjust-smoking_6-27-2022.xlsx"))
logistic_mixed_model_ad_cn.females.SSc <- glmm_female_result
logistic_mixed_model_ad_cn.females.SSc$file_order <- NULL
readr::write_csv(
  logistic_mixed_model_ad_cn.females.SSc, 
  file = file.path(dir.results.glmm, "Females_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_smoking.csv")
)
```


### Adjust for education years

#### Male 

```{r eval = F}
glmm_male_result <- read_xlsx(
  file.path(dir.base.data2,"LW/Males_SAS_adjust_education/RESULTS/ad-vs-cn-males-adjust-education_6-29-2022.xlsx"))
logistic_mixed_model_ad_cn.males.edu <- glmm_male_result
logistic_mixed_model_ad_cn.males.edu$file_order <- NULL
readr::write_csv(
  logistic_mixed_model_ad_cn.males.edu, 
  file = file.path(dir.results.glmm, "Males_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_edu.csv")
)
```

#### Female 

```{r eval = F}
glmm_female_result <- read_xlsx(
  file.path(dir.base.data2,"LW/Females_SAS_adjust_education/results/ad-vs-cn-females-adjust-education_6-29-2022.xlsx"))
logistic_mixed_model_ad_cn.females.edu <- glmm_female_result
logistic_mixed_model_ad_cn.females.edu$file_order <- NULL
readr::write_csv(
  logistic_mixed_model_ad_cn.females.edu, 
  file = file.path(dir.results.glmm, "Females_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_edu.csv")
)
```


## Annotate results
```{R annotate_functions, eval = FALSE}
message("o Stating to annotate")
adni.se <- readRDS(file.path(dir.data.processed, "adni_se_min_age_at_visit_65_with_XY_AD_CN.RDS"))
load(file.path(dir.data.aux,"great_EPIC_array_annotation.rda"))
load(file.path(dir.data.aux,"E073_15_coreMarks_segments.rda"))
load(file.path(dir.data.aux,"meta_analysis_cpgs.rda"))

annotate_results <- function(result, data.se){
  #result <- result[result$cpg %in% rownames(data.se),]
  result$fdr <- p.adjust(result$Probt,method = "fdr")
  result <- cbind(
    result,
    rowRanges(data.se)[result$cpg] %>% as.data.frame() %>% dplyr::select(1:4)
  )
  result$Islands.UCSC.Relation_to_Island <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Islands.UCSC[result$cpg,"Relation_to_Island"]
  result$UCSC_RefGene_Name <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Other[result$cpg,"UCSC_RefGene_Name"]       
  result$UCSC_RefGene_Group <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Other[result$cpg,"UCSC_RefGene_Group"]     
  
  result <- dplyr::left_join(result, great,by = c("seqnames","start","end","cpg"))
  
  hits <- findOverlaps(rowRanges(data.se)[result$cpg],ChmmModels.gr) %>% as.data.frame()
  hits$state <- ChmmModels.gr$state[hits$subjectHits]
  hits$cpg <- result$cpg[hits$queryHits]
  result <- merge(result, hits[,c("state","cpg")], all.x = TRUE, sort = FALSE, by = "cpg")
  
  result$sig.in.brain <- result$cpg %in% brain.meta.analysis.cpgs
  return(result)
}
```

```{R annotate, eval = FALSE}
## No adjust smoking scores
logistic_mixed_model_ad_cn.females <- readr::read_csv( 
  file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_SAS.csv")
)

logistic_mixed_model_ad_cn.females <- logistic_mixed_model_ad_cn.females %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.females,
  file = file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_all_annotated.csv"),
)

logistic_mixed_model_ad_cn.males <- readr::read_csv( 
  file.path(dir.results.glm,"Males_ADNI_AD_vs_CN_glmm_using_beta_SAS.cvs")
)

logistic_mixed_model_ad_cn.annotated.males <- logistic_mixed_model_ad_cn.males %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.annotated.males,
  file = file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_all_annotated.csv"),
)
## Adjust smoking scores
logistic_mixed_model_ad_cn.females.SSc <- readr::read_csv( 
  file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_smoking.csv")
)

logistic_mixed_model_ad_cn.annotated.females.SSc <- logistic_mixed_model_ad_cn.females.SSc %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.annotated.females.SSc,
  file = file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_smoking_all_annotated.csv"),
)

logistic_mixed_model_ad_cn.males.SSc <- readr::read_csv( 
  file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_smoking.csv")
)

logistic_mixed_model_ad_cn.annotated.males.SSc <- logistic_mixed_model_ad_cn.males.SSc %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.annotated.males.SSc,
  file = file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_smoking_all_annotated.csv"),
)
## Adjust education year
logistic_mixed_model_ad_cn.females.edu <- readr::read_csv( 
  file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_edu.csv")
)

logistic_mixed_model_ad_cn.annotated.females.edu <- logistic_mixed_model_ad_cn.females.edu %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.annotated.females.edu,
  file = file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_all_annotated.csv"),
)

logistic_mixed_model_ad_cn.males.edu <- readr::read_csv( 
  file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_SAS_adjust_edu.csv")
)

logistic_mixed_model_ad_cn.annotated.males.edu <- logistic_mixed_model_ad_cn.males.edu %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.annotated.males.edu,
  file = file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_all_annotated.csv"),
)
```

```{R read_results, include = FALSE, eval = T}
logistic_mixed_model_ad_cn.annotated.females <- readr::read_csv(
  file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_all_annotated.csv")
)
logistic_mixed_model_ad_cn.annotated.males <- readr::read_csv(
  file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_all_annotated.csv")
)
logistic_mixed_model_ad_cn.annotated.females.SSc <- readr::read_csv(
  file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_smoking_all_annotated.csv")
)
logistic_mixed_model_ad_cn.annotated.males.SSc <- readr::read_csv(
  file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_smoking_all_annotated.csv")
)
logistic_mixed_model_ad_cn.annotated.females.edu <- readr::read_csv(
  file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_all_annotated.csv")
)
logistic_mixed_model_ad_cn.annotated.males.edu <- readr::read_csv(
  file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_all_annotated.csv")
)
```

# Calculate inflation factors and Bacon correction for final results

## Auxiliary functions

```{R inflation, message = FALSE, results = 'hide', eval = T}
library(dplyr)
library(bacon)
library(GWASTools)

estimation_of_inflation <- function(data){
  ### 1. Compute genomic inflation factor before bacon adjustment
  data$zvalue <- data$Estimate / data$StdErr
  #data$zvalue <- qnorm(pt(data$tvalue, df))
  data$chisq <- (data$zvalue) ^ 2
  # inflation factor - last term is median from chisq distrn with 1 df  
  inflationFactor <- median(data$chisq,na.rm = TRUE) / qchisq(0.5, 1)
  print("o Before bacon correction")
  print("Conventional lambda")
  print(inflationFactor)
  ### 2. bacon analysis
  bc <- bacon(
    teststatistics = NULL,
    effectsizes =  data$Estimate,
    standarderrors = data$StdErr,
    na.exclude = TRUE
  )
  # inflation factor
  print("bacon lambda")
  print(inflation(bc))
   print("bias")
  print(bias(bc))
  
  ### 3. Create final dataset
  data.with.inflation <- data.frame(
    data,
    Estimate.bacon = bacon::es(bc),
    StdErr.bacon = bacon::se(bc),
    pValue.bacon = pval(bc),
    fdr.bacon = p.adjust(pval(bc), method = "fdr"),
    stringsAsFactors = FALSE
  )
  print("o After bacon correction")
  print("Conventional lambda")
  print(median((data.with.inflation$Estimate.bacon/data.with.inflation$StdErr.bacon) ^ 2,na.rm = TRUE)/qchisq(0.5, 1))
  print("bacon lambda")
  print(estimates(bc))
  
  percent_null <- trunc ( estimates(bc)[1]*100, digits = 0)
  percent_1  <- trunc ( estimates(bc)[2]*100, digits = 0 )
  percent_2  <- 100 - percent_null - percent_1  
  bc2 <- bacon(
    teststatistics = NULL,
    effectsizes =  data.with.inflation$Estimate.bacon,
    standarderrors = data.with.inflation$StdErr.bacon,
    na.exclude = TRUE, 
    priors = list(sigma = list(alpha = 1.28, beta =
                                 0.36), mu = list(lambda = c(0, 3, -3), tau = c(1000, 100, 100)), epsilon
                  = list(gamma = c(percent_null, percent_1, percent_2)))
  )
  
  print(inflation(bc2))
  print(estimates(bc2))
  
  data.with.inflation <- data.with.inflation %>% dplyr::select(-c(zvalue, chisq))
  return(
    list(
      "data.with.inflation" = data.with.inflation,
      "inflationFactor" =   inflationFactor,
      "estimatedInflation" = inflation(bc),
      "baconObject" = bc
    )
  )
}


plot_inflation <- function(data, main){
  qqPlot(
    data$data.with.inflation$Probt, ci = TRUE, 
    main = paste0(main)
  ) + text(
    x = 0.5, 
    y = -log10(min(data$data.with.inflation$Probt)) - 0.5,
    labels = bquote(lambda == .(format(data$inflationFactor,digits = 4))),
    pos = 4
  ) + text(
    x = 0.5, 
    y = -log10(min(data$data.with.inflation$Probt)) - 1,
    labels = bquote(lambda[bacon] == .(format(data$estimatedInflation,digits = 4))),
    pos = 4
  )
}
```

### Males

#### No adjust smoking scores

```{R, eval = T}
logistic_mixed_model_ad_cn.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated.males)
dim(logistic_mixed_model_ad_cn.with.inflation$data.with.inflation)
```


```{r}
plot(logistic_mixed_model_ad_cn.with.inflation$baconObject) 
```



```{r}
fit(logistic_mixed_model_ad_cn.with.inflation$baconObject, n=100)
```


```{r eval = F, echo = F}
ggsave(file.path(dir.base.data2,"analysis_results/", "others/distribution_plot/ADNI_Male_test_stat_and_null_plot.pdf"),
       width = 12,
       height = 9 )
```


```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.with.inflation$data.with.inflation,
  file = file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_with_inflation.csv")
)
```



```{R, fig.width = 6, eval = T}
plot_inflation(
  data = logistic_mixed_model_ad_cn.with.inflation, 
  main = "MALES ADNI AD vs CN - single cpg\n Logistic regression model"
)
```

```{r echo = F}
rm(logistic_mixed_model_ad_cn.with.inflation)
```


#### Adjust smoking scores

```{R, eval = T}
logistic_mixed_model_ad_cn.SSc.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated.males.SSc)
dim(logistic_mixed_model_ad_cn.SSc.with.inflation$data.with.inflation)
```

```{r}
plot(logistic_mixed_model_ad_cn.SSc.with.inflation$baconObject)
```

```{r eval = F, echo = F}
ggsave(file.path(dir.base.data2,"analysis_results/",  "others/distribution_plot/ADNI_Male_adjust_SSc_test_stat_and_null_plot.pdf"),
       width = 12,
       height = 9 )
```

```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.SSc.with.inflation$data.with.inflation,
  file = file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_smoking_with_inflation.csv")
)
```



```{R, fig.width = 6, eval = T}
plot_inflation(
  data = logistic_mixed_model_ad_cn.SSc.with.inflation, 
  main = "MALES ADNI AD vs CN - single cpg\n Logistic regression model"
)
```



```{r echo = F}
rm(logistic_mixed_model_ad_cn.SSc.with.inflation)
```

#### Adjust education year

```{r}
logistic_mixed_model_ad_cn.edu.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated.males.edu)
dim(logistic_mixed_model_ad_cn.edu.with.inflation$data.with.inflation)
```

```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.edu.with.inflation$data.with.inflation,
  file = file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_with_inflation.csv")
)
```

```{R, fig.width = 6, eval = T}
plot_inflation(
  data = logistic_mixed_model_ad_cn.edu.with.inflation, 
  main = "MALE - ADNI AD vs CN - single cpg\n Logistic regression model"
)
```

```{r}
rm(logistic_mixed_model_ad_cn.edu.with.inflation)
```



### Females

#### No adjust smoking scores

```{R, eval = T}
logistic_mixed_model_ad_cn.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated.females)
dim(logistic_mixed_model_ad_cn.with.inflation$data.with.inflation)
```

```{r}
plot(logistic_mixed_model_ad_cn.with.inflation$baconObject)
```

```{r eval = F, echo = F}
ggsave(file.path(dir.base.data2,"analysis_results/", "others/distribution_plot/ADNI_Female_test_stat_and_null_plot.pdf"),
       width = 12,
       height = 9 )
```

```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.with.inflation$data.with.inflation,
  file = file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_with_inflation.csv")
)
```

```{R, fig.width = 6, eval = T}
plot_inflation(
  data = logistic_mixed_model_ad_cn.with.inflation, 
  main = "FEMALE - ADNI AD vs CN - single cpg\n Logistic regression model"
)
```


```{r echo = F}
rm(logistic_mixed_model_ad_cn.with.inflation)
```

#### Adjust smoking scores

```{r}
logistic_mixed_model_ad_cn.SSc.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated.females.SSc)
dim(logistic_mixed_model_ad_cn.SSc.with.inflation$data.with.inflation)
```

```{r}
plot(logistic_mixed_model_ad_cn.SSc.with.inflation$baconObject)
```

```{r eval = F, echo = F}
ggsave(file.path(dir.base.data2,"analysis_results/", "others/distribution_plot/ADNI_Female_adjust_SSc_test_stat_and_null_plot.pdf"),
       width = 12,
       height = 9 )
```

```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.SSc.with.inflation$data.with.inflation,
  file = file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_smoking_with_inflation.csv")
)
```

```{R, fig.width = 6, eval = T}
plot_inflation(
  data = logistic_mixed_model_ad_cn.SSc.with.inflation, 
  main = "FEMALE - ADNI AD vs CN - single cpg\n Logistic regression model"
)
```

```{r echo = F}
rm(logistic_mixed_model_ad_cn.SSc.with.inflation)
```


#### Adjust education year

```{r}
logistic_mixed_model_ad_cn.edu.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated.females.edu)
dim(logistic_mixed_model_ad_cn.edu.with.inflation$data.with.inflation)
```

```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.edu.with.inflation$data.with.inflation,
  file = file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_with_inflation.csv")
)
```

```{R, fig.width = 6, eval = T}
plot_inflation(
  data = logistic_mixed_model_ad_cn.edu.with.inflation, 
  main = "FEMALE - ADNI AD vs CN - single cpg\n Logistic regression model"
)
```

```{r echo = F}
rm(logistic_mixed_model_ad_cn.edu.with.inflation)
```



# Correlation and test

## No adjust education year vs adjust education year

### Male


```{r}
Male.logistic_mixed_model_ad_cn.with.inflation <- readr::read_csv(
    file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_with_inflation.csv")
)
Male.logistic_mixed_model_ad_cn.with.inflation <- Male.logistic_mixed_model_ad_cn.with.inflation %>%
  dplyr::select(
    cpg, Estimate.bacon, pValue.bacon, fdr.bacon 
  )
Male.logistic_mixed_model_ad_cn.edu.with.inflation <- readr::read_csv(
    file.path(dir.results.glmm,"Males_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_with_inflation.csv")
)
Male.logistic_mixed_model_ad_cn.edu.with.inflation <- Male.logistic_mixed_model_ad_cn.edu.with.inflation %>%
  dplyr::select(
    cpg, Estimate.bacon, pValue.bacon, fdr.bacon 
  )
colnames(Male.logistic_mixed_model_ad_cn.edu.with.inflation)[-1] <- 
  paste0("adjustEDU_", colnames(Male.logistic_mixed_model_ad_cn.edu.with.inflation)[-1])
merge_df <- left_join(
  Male.logistic_mixed_model_ad_cn.with.inflation,
  Male.logistic_mixed_model_ad_cn.edu.with.inflation
)
merge_df.sig <- merge_df %>% filter(
  pValue.bacon < 1E-05 | adjustEDU_pValue.bacon < 1E-05
)
```

#### Coefficients


```{r}
### all coefficients
cor.test(
  merge_df$Estimate.bacon, 
  merge_df$adjustEDU_Estimate.bacon,
  method = "spearman"
)
### significant coefficients
cor.test(
  merge_df.sig$Estimate.bacon, 
  merge_df.sig$adjustEDU_Estimate.bacon,
  method = "spearman"
)
```


```{r echo = F}
ggscatter(merge_df,
          x = "Estimate.bacon", y = "adjustEDU_Estimate.bacon",
          size = 0.1) + 
  stat_cor(method = "spearman", label.x = -1, label.y = 1) + 
  xlab("Coefficients from No adjust EDU year") + 
  ylab("Coefficients from Adjust EDU year") + 
  ggtitle("ADNI Male coefficients of all CpG")
```


```{r echo = F}
ggscatter(merge_df.sig,
          x = "Estimate.bacon", y = "adjustEDU_Estimate.bacon",
          size = 1) + 
  stat_cor(method = "spearman", label.x = -0.4, label.y = 0.5) + 
  geom_text_repel(aes(label = cpg), size = 2) +
  xlab("Coefficients from No adjust EDU year") + 
  ylab("Coefficients from Adjust EDU year") + 
  ggtitle("ADNI Male coefficients of significant CpG (pval < 1E-05)")
```

```{r}
ggsave(
  file.path(dir.results, "MALE_ADNI_Adjust_edu_year_vs_ADNI_NO_adjust_edu_year.pdf"),
  width = 8, 
  height = 6
)
```


#### pValues


```{r}
### all pvalues
cor.test(
  merge_df$pValue.bacon, 
  merge_df$adjustEDU_pValue.bacon,
  method = "spearman"
)
### significant pvalues
cor.test(
  merge_df.sig$pValue.bacon, 
  merge_df.sig$adjustEDU_pValue.bacon,
  method = "spearman"
)
```


```{r echo = F}
ggplot(merge_df, 
       aes(x = -log10(pValue.bacon), y = -log10(adjustEDU_pValue.bacon))) + 
  geom_point(size = 0.5) + theme_classic() +
  stat_cor(method = "spearman", label.x = 0, label.y = 4) + 
  xlab("-log10(pValues) from No adjust EDU year") + 
  ylab("-log10(pValues) from adjust EDU year") + 
  ggtitle("ADNI Male all CpGs pValues scatter plot")
```



```{r echo = F}
ggplot(merge_df.sig, 
       aes(x = -log10(pValue.bacon), y = -log10(adjustEDU_pValue.bacon))) + 
  geom_point(size = 1) + theme_classic() +
  stat_cor(method = "spearman", label.x = 0, label.y = 4) + 
  geom_text_repel(aes(label = cpg), size = 2) +
  xlab("-log10(pValues) from No adjust EDU year") + 
  ylab("-log10(pValues) from adjust EDU year") + 
  ggtitle("ADNI Male significant CpGs pValues scatter plot")
```


### Female

```{r}
Female.logistic_mixed_model_ad_cn.with.inflation <- readr::read_csv(
    file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_with_inflation.csv")
)
Female.logistic_mixed_model_ad_cn.with.inflation <- Female.logistic_mixed_model_ad_cn.with.inflation %>%
  dplyr::select(
    cpg, Estimate.bacon, pValue.bacon, fdr.bacon 
  )
Female.logistic_mixed_model_ad_cn.edu.with.inflation <- readr::read_csv(
    file.path(dir.results.glmm,"Females_ADNI_AD_vs_CN_glmm_using_beta_adjust_edu_with_inflation.csv")
)
Female.logistic_mixed_model_ad_cn.edu.with.inflation <- Female.logistic_mixed_model_ad_cn.edu.with.inflation %>%
  dplyr::select(
    cpg, Estimate.bacon, pValue.bacon, fdr.bacon 
  )
colnames(Female.logistic_mixed_model_ad_cn.edu.with.inflation)[-1] <- 
  paste0("adjustEDU_", colnames(Female.logistic_mixed_model_ad_cn.edu.with.inflation)[-1])
merge_df <- left_join(
  Female.logistic_mixed_model_ad_cn.with.inflation,
  Female.logistic_mixed_model_ad_cn.edu.with.inflation
)
merge_df.sig <- merge_df %>% filter(
  pValue.bacon < 1E-05 | adjustEDU_pValue.bacon < 1E-05
)
```

#### Coefficients


```{r}
### all coefficients
cor.test(
  merge_df$Estimate.bacon, 
  merge_df$adjustEDU_Estimate.bacon,
  method = "spearman"
)
### significant coefficients
cor.test(
  merge_df.sig$Estimate.bacon, 
  merge_df.sig$adjustEDU_Estimate.bacon,
  method = "spearman"
)
```


```{r echo = F}
ggscatter(merge_df,
          x = "Estimate.bacon", y = "adjustEDU_Estimate.bacon",
          size = 0.1) + 
  stat_cor(method = "spearman", label.x = -1, label.y = 1) + 
  xlab("Coefficients from No adjust EDU year") + 
  ylab("Coefficients from Adjust EDU year") + 
  ggtitle("ADNI Female coefficients of all CpG")
```

```{r echo = F}
ggscatter(merge_df.sig,
          x = "Estimate.bacon", y = "adjustEDU_Estimate.bacon",
          size = 1) + 
  stat_cor(method = "spearman", label.x = -0.4, label.y = 0.5) + 
  geom_text_repel(aes(label = cpg), size = 2) +
  xlab("Coefficients from No adjust EDU year") + 
  ylab("Coefficients from Adjust EDU year") + 
  ggtitle("ADNI Female coefficients of significant CpG (pval < 1E-05)")
```

```{r}
ggsave(
  file.path(dir.results, "FEMALE_ADNI_Adjust_edu_year_vs_ADNI_NO_adjust_edu_year.pdf"),
  width = 8, 
  height = 6
)
```

#### pValues


```{r}
### all pvalues
cor.test(
  merge_df$pValue.bacon, 
  merge_df$adjustEDU_pValue.bacon,
  method = "spearman"
)
### significant pvalues
cor.test(
  merge_df.sig$pValue.bacon, 
  merge_df.sig$adjustEDU_pValue.bacon,
  method = "spearman"
)
```


```{r echo = F}
ggplot(merge_df, 
       aes(x = -log10(pValue.bacon), y = -log10(adjustEDU_pValue.bacon))) + 
  geom_point(size = 0.5) + theme_classic() +
  stat_cor(method = "spearman", label.x = 0, label.y = 4) + 
  xlab("-log10(pValues) from No adjust EDU year") + 
  ylab("-log10(pValues) from adjust EDU year") + 
  ggtitle("ADNI Female all CpGs pValues scatter plot")
```


```{r echo = F}
ggplot(merge_df.sig, 
       aes(x = -log10(pValue.bacon), y = -log10(adjustEDU_pValue.bacon))) + 
  geom_point(size = 1) + theme_classic() +
  stat_cor(method = "spearman", label.x = 0, label.y = 4) + 
  geom_text_repel(aes(label = cpg), size = 2) +
  xlab("-log10(pValues) from No adjust EDU year") + 
  ylab("-log10(pValues) from adjust EDU year") + 
  ggtitle("ADNI Female significant CpGs pValues scatter plot")
```

## Education year vs AD Wilcoxon test


```{R}
pheno.cn.ad <- readr::read_csv(file.path(dir.data.processed,"ADNI_Se_colData_with_XY_AD_CN.csv"))
pheno.cn.ad.last_visit <- pheno.cn.ad %>% 
  group_by(RID) %>%
  dplyr::slice(which.max(as.Date(DateDrawn, '%m/%d/%Y'))) %>%
  mutate(DX0 = ifelse(DX == "Dementia", "AD", "CN"))
pheno.cn.ad.first_visit <- pheno.cn.ad %>% 
  group_by(RID) %>%
  dplyr::slice(which.min(as.Date(DateDrawn, '%m/%d/%Y'))) 
```


```{r}
## Male
pheno.cn.ad.last_visit %>% filter(PTGENDER == 'Male') %>%
  wilcox.test(
    PTEDUCAT ~ DX,
    data = .
  )
## Female
pheno.cn.ad.last_visit %>% filter(PTGENDER == 'Female') %>%
  wilcox.test(
    PTEDUCAT ~ DX,
    data = .
  )
```

```{r}
## Male
pheno.cn.ad.first_visit %>% filter(PTGENDER == 'Male') %>%
  wilcox.test(
    PTEDUCAT ~ DX,
    data = .
  )
## Female
pheno.cn.ad.first_visit %>% filter(PTGENDER == 'Female') %>%
  wilcox.test(
    PTEDUCAT ~ DX,
    data = .
  )
```


```{r}
ggboxplot(pheno.cn.ad.last_visit, x = "DX0", y = "PTEDUCAT",
          facet.by = "PTGENDER", fill = "DX0", palette = "Blues",
          title = "ADNI last visit education year vs AD")  + labs(fill = "DX", x = "DX") + 
  geom_jitter(shape=16, size = 0.8, position=position_jitter(0.2)) +
  stat_compare_means(method = 'wilcox.test', label.y.npc = 'bottom')
```

```{r eval = F}
ggsave(
  file.path(dir.results, "ADNI_last_visit_edu_year_vs_AD.pdf"),
  width = 10, 
  height = 6
)
```


```{r}
ggboxplot(pheno.cn.ad.first_visit, x = "DX", y = "PTEDUCAT",
          facet.by = "PTGENDER", fill = "DX", palette = "Blues",
          title = "ADNI first visit education year vs AD") +
  geom_jitter(shape=16, size = 0.8, position=position_jitter(0.2)) +
  stat_compare_means(method = 'wilcox.test', label.y.npc = 'bottom')
```

# Session information
```{R}
devtools::session_info()
```